---
title: Segmenting Medicare STAR measures based on time-based trends in the distributions of contract scores
author: Danny Morris
date: "`r Sys.Date()`"
output: 
  blogdown::html_page:
    toc: true
    highlight: pygments
slug: clustering-medicare-star-measures-using-time-based-statistics
categories:
  - R
  - Analytics
  - Cluster Analysis
tags:
  - R
  - Analytics
  - Cluster Analysis
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

## Overview

This post demonstrates the use of hierarchical clustering to segment Medicare STAR measures based on trends in the distribution of contract scores. Trends are quantified using three time-based statistics: 1) slope of the average score (positive slope suggests scores are improving across the board), 2) slope of the variance (negative slope suggests scores are becoming more tightly distributed across the board), and 3) distance to capacity (high value suggests scores are far from their maximum value and have room for improvement). Understanding trends within each measure enables data-driven planning and goal setting.

## Packages

```{r}
library(tidyverse)
library(gridExtra)
library(RColorBrewer)
library(ggdendro)
library(ggiraph)
```

## Data

The data set is publically available at the [CMS website](https://www.cms.gov/Medicare/Prescription-Drug-Coverage/PrescriptionDrugCovGenIn/PerformanceData). Download the cleansed version [here](.files/standardized_scores.csv).

```{r}
scores_df <- read_csv("standardized_scores.csv")
```

## Sample measures

This analysis is limited to current Part C measures that are scored as a rate between 0 and 1.

```{r}
sample_measures <- scores_df %>%
  filter(part == "C") %>%
  group_by(clean_measure) %>%
  summarise(
    n_years = n_distinct(star_year),
    most_recent_year = max(star_year),
    min_score = min(score, na.rm = T),
    max_score = max(score, na.rm = T)
  ) %>%
  filter(most_recent_year == 2020,
         n_years >= 3, 
         min_score >= 0,
         max_score <= 1) %>%
  select(clean_measure) %>%
  distinct()

sample_measures %>%
  select(clean_measure) %>%
  pull()
```

## Contract scores

```{r}
scores <- scores_df %>%
  inner_join(sample_measures, by = "clean_measure") %>%
  select(clean_measure,
         CONTRACT_ID,
         star_year,
         score) %>%
  drop_na()

scores
```

## Defining the relevant time-based statistics

The attributes of interest include:

1. **Improvement trend**

This statistic reflects the slope of the average contract score. It answers the question "are the contract scores trending upward or downward?" A highly positive slope suggests scores are rapidly improving across the board. 

2. **Variance trend**

This statistic reflects the slope of the contract score varaince. It answers the question "are the contract becoming wider or narrower?" A highly negative slope means the variance is decreasing, suggesting contracts are becoming more competitive.

3. **Capacity**

This statistic reflects the distance between the average score and the maximum score. For the measures in this analysis, the maximum score is 1. This statistic answers the question "on average, how much room is available for the contract scores to grow?" High values suggest plenty of room for the contract scores to improve. Low values suggest the contract scores are hovering near the maximum value with little room for growth.

These attributes can be quantified and used in a clustering model to reveal interesing segments of Medicare STAR measures.

## Calculate and visualize attributes

```{r}
features <- scores %>% 
  split(.$clean_measure) %>%
  map(., function(measure) {
    year_summary <- measure %>%
      drop_na() %>%
      group_by(clean_measure, star_year) %>%
      summarise(
        yr_n_contracts = n_distinct(CONTRACT_ID),
        yr_median_score = median(score),
        yr_mad_score = mad(score),
        yr_capacity = 1 - yr_median_score
      ) %>%
      ungroup() %>%
      mutate(year_id = seq_along(star_year))
    
    year_summary %>%
      group_by(clean_measure) %>%
      summarise(
        improvement_trend = lm(yr_median_score ~ year_id)$coefficients[2],
        variance_trend = lm(yr_mad_score ~ year_id)$coefficients[2],
        capacity = mean(yr_capacity)
      )
  }) %>%
  bind_rows()
```

```{r}
p <- features %>%
  ggplot(aes(x = improvement_trend, y = variance_trend)) +
  geom_point_interactive(aes(size = capacity,
                             tooltip = paste(clean_measure)),
                         alpha = 0.8) +
  geom_vline(aes(xintercept = 0), 
             color = 'gray50') +
  geom_hline(aes(yintercept = 0), 
             color = 'gray50') +
  labs(title = "Medicare STAR measure attributes",
       x = "Improvement Trend",
       y = "Variance Trend",
       size = "Capacity Trend") +
  theme_bw()

ggiraph::girafe(code = print(p), 
                height_svg = 4)
```

## Apply hierarchical clustering

### Center and scale variables using median and median absolute deviation

```{r}
robust_scale <- function(x) {
  (x - median(x)) / mad(x)
}

model_df <- features %>%
  select(-clean_measure) %>%
  mutate_all(list(robust_scale))
```

### Clustering

```{r, fig.height = 9}
set.seed(1234)

hier_clustering <- model_df %>%
  dist(method = "euclidean") %>%
  hclust(method = "ward.D2")

hier_clustering$labels <- features$clean_measure

ggdendro::ggdendrogram(hier_clustering, rotate = T) +
  labs(title = "Clustering dendrogram")
```

Since this analysis is exploratory, there is no correct answer to the number of clusters present in the data. Based on a visual interpretation of the dendrogram, between 4 and 6 clusters is reasonable. 

```{r}
clusters <- cutree(hier_clustering, k = 6)
```

### Visualize clusters

```{r, fig.height=9}
cluster_ids <- seq(1, max(clusters), 1)
colors <- RColorBrewer::brewer.pal(max(clusters), "Set1")

plist <- map2(cluster_ids, colors, function(id, color) {
  features %>%
    mutate(Cluster = clusters) %>%
    mutate(target_cluster = ifelse(Cluster == id, "1", "0")) %>%
    ggplot(aes(x = improvement_trend, y = variance_trend)) +
    geom_point(aes(color = target_cluster, size = capacity)) +
    scale_color_manual(values = c("gray80", color)) +
    labs(title = paste("Cluster:", id)) +
    guides(color = FALSE) +
    theme_bw() 
})

do.call("grid.arrange", c(plist, ncol = floor(sqrt(length(plist)))))
```

## Conclusion

The measure attributes calculated in the analysis reveal interesting segments of Medicare STAR measures based on trends in performance, variance, and capacity. For example, Cluster 1 contains measures that are holding steady in their distributions of contract scores. Cluster 5 contains a single outlier measure with contract scores that are trending upwards at a rapid rate and becoming narrow, suggesting health plans are rapidly improving in this measure and becoming more competitive.

Because there are many Medicare STAR measures, it can be difficult to assess each measure's individual behavior. An alternative approach is to quantify meaningful measure attributes and discover segments using clustering methods. This technique facilitates an immediate understanding of the current landscape of Medicare STAR measures and their trends without having to delve into each measure individually. It also provides an immediate basis for discussion and goal setting.






















