---
title: Quantifying "month composition" to improve monthly forecasts (Python)
author: Danny Morris
date: '2020-06-17'
output: 
  blogdown::html_page:
    toc: true
    highlight: pygments
slug: applying-calendar-adjustments-to-improve-monthly-health-insurance-claims-forecasts-python
categories:
  - Python
  - Forecasting
tags:
  - Python
  - Forecasting
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

## Overview

This post is essentialy a copy of [this post]((/2020/06/17/monthly-forecasts-using-effective-days-r/)) except the methodology is written in Python using pandas, numpy, and Facebook Prophet.

```{r, echo=F}
# Configure conda env
Sys.setenv(
  RETICULATE_PYTHON = "C:/Users/Owner/.conda/envs/dsci/python.exe"
)

reticulate::use_condaenv("dsci")
```

```{python, echo=F}
import os
os.environ['QT_QPA_PLATFORM_PLUGIN_PATH'] = 'C:/Users/Owner/.conda/envs/dsci/Library/plugins/platforms'
```

## Python packages

```{python}
import pandas as pd # version 1.0.5
import numpy as np  # version 1.18.5
import seaborn as sns
import matplotlib
import matplotlib.pyplot as plt
import fbprophet
from fbprophet import Prophet
```

```{python}
print("Pandas version: " + pd.__version__)
print("Numpy version: " + np.__version__)
print("Seaborn version: " + sns.__version__)
print("Matplotlib version: " + matplotlib.__version__)
print("Prophet version: " + fbprophet.__version__)
```

## Data

### Daily claims

```{python}
claims = pd.read_csv('sample_claims.csv', parse_dates=['Inc_Month'])

claims.head()
```

### Monthly claims

```{python}
monthly_claims = (claims
   .groupby(['Inc_Year', 'Inc_Month'])
   .agg(Claims = ('Claims', 'sum'),
        Population = ('Population', 'mean'))
   .reset_index()
   )

ax = sns.lineplot(data=monthly_claims, x='Inc_Month', y='Claims')
ax.set_title("Daily claims")
```

## Quantify month composition

### Weekday occurences per month

```{python}
weekday_counts = (claims
   .groupby(['Inc_Month', 'Inc_Year', 'DOW_Name', 'DOW'])
   .agg(Total_Claims = ('Claims', 'sum'),
        Occurences = ('Date', 'nunique'))
   .reset_index()
 )
```

### Weekday relative values per month

```{python}
ax = sns.boxplot(x="DOW", y="Claims", data=claims)
ax.set_title('Claims distribution by day of week')
```

```{python}
weekday_averages = (weekday_counts
   .assign(Weekday_Average = weekday_counts['Total_Claims']/weekday_counts['Occurences'])
 )

average_weekday = (weekday_averages
   .groupby(['Inc_Month'])
   .agg(Average_Weekday = ('Weekday_Average', 'mean'))
   .reset_index()
 )

weekday_relatives = (weekday_averages.merge(average_weekday,
                       on = 'Inc_Month',
                       how = 'inner')
   .pipe(lambda x: x.assign(DOW_Ratio = x['Weekday_Average']/x['Average_Weekday']))
   .filter(['Inc_Month',
            'Weekday_Average',
            'DOW',
            'DOW_Name',
            'DOW_Ratio'])
 )

ax = sns.FacetGrid(weekday_relatives, col="DOW", 
                  col_wrap=3, sharey=False)
ax.map(sns.lineplot, "Inc_Month", "DOW_Ratio", color=".3")
plt.show()
plt.close()
```

### Calcualte effective days

```{python}
effective_days = (weekday_counts.merge(weekday_relatives,
                      on = ['Inc_Month', 'DOW_Name'],
                      how = 'inner')
   .pipe(lambda x: x.assign(Effective_Days = x['Occurences']*x['DOW_Ratio']))
   .groupby(['Inc_Month', 'Inc_Year'])
   .agg(Actual_Days = ('Occurences', 'sum'),
        Effective_Days = ('Effective_Days', 'sum'))
   .reset_index()
 )

average_effective_days = (
    effective_days
      .groupby(['Inc_Year'])
      .agg(Avg_Effective_Days = ('Effective_Days', 'mean'))
      .reset_index()
 )

effective_days_index = (effective_days.merge(average_effective_days,
                     on = 'Inc_Year',
                     how = 'inner')
   .pipe(lambda x: x.assign(Effective_Days_Index = x['Effective_Days']/x['Avg_Effective_Days']))
   .filter(['Inc_Month',
            'Inc_Year',
            'Effective_Days',
            'Effective_Days_Index'])
 )
```

## Calculate adjusted PMPM using month composition and population adjustments

```{python}
pmpm = (monthly_claims
   .merge(effective_days_index, on = "Inc_Month", how = "inner")
   .pipe(lambda x: x.assign(Adj_PMPM = x['Claims']/x['Population']/x['Effective_Days_Index']))
   .pipe(lambda x: x.assign(Raw_PMPM = x['Claims']/x['Population']))
   .filter(['Inc_Month', 
            'Adj_PMPM', 
            'Raw_PMPM', 
            'Claims', 
            'Population',
            'Effective_Days_Index'])
 )

pmpm.head()
```

## Compare RMSE of forecasts using adjusted and unadjusted PMPM

The algorithm chosen for this comparison is Facebook Prophet.

```{python}
def calc_rmse(adjusted = True):

    '''
    Train model using adjusted or unadjusted PMPM as target
    Returns RMSE on testing set
    '''
    
    # define target
    if adjusted:
        target = "Adj_PMPM"
    else:
        target = "Raw_PMPM"

    # train/test splits
    training = (pmpm[pmpm['Inc_Month'] <= '2018-12-01']
       .rename(columns={"Inc_Month": "ds", target: "y"})
    )
    
    testing = (pmpm[pmpm['Inc_Month'] >= '2019-01-01']
       .rename(columns={"Inc_Month": "ds", target: "y"})
       .reset_index()
    )
    
    # fit Prophet
    m = Prophet()
    m.fit(training)
    
    # make forecasts
    forecast = m.predict(testing)
    
    # function to convert PMPM to original claims units
    def backtransform(x, adjusted):
        if adjusted:
            out = x['yhat']*x['Population']*x['Effective_Days_Index']
        else:
            out = x['yhat']*x['Population']
        return(out)   
    
    # calculate forecast errors
    forecast_errors = (pd.concat(
      [forecast,
       testing[['Population', 'Effective_Days_Index', 'Claims']]],
       axis=1)
     .pipe(lambda x: x.assign(Forecast = backtransform(x, adjusted)))
     .pipe(lambda x: x.assign(Error = x['Forecast']-x['Claims']))
    )

    # calculate RMSE
    mse = np.mean(forecast_errors['Error']**2)
    rmse = np.sqrt(mse)
    
    return(rmse)
```

```{python}
rmse_w_adj = calc_rmse(adjusted=True)
rmse_wo_adj = calc_rmse(adjusted=False)

plt_X = ['Adjusted PMPM', 'Unadjusted PMPM']
plt_Y = [rmse_w_adj, rmse_wo_adj]
plt.bar(plt_X, plt_Y, color = ['blue', 'gray'])
plt.title("Adjusting PMPM using month composition improves RMSE")
plt.show()
```
