---
title: Quantifying "month composition" to improve monthly forecasts
author: Danny Morris
date: "2020-06-17"
output: 
  blogdown::html_page:
    toc: true
    highlight: pygments
slug: monthly-forecasts-using-effective-days-r
categories:
  - Forecasting
  - R
  - Predictive Modeling
tags:
  - R
  - Forecasting
  - Predictive Modeling
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

## Overview

This post describes a simple yet effective technique for improving the accuracy of monthly forecasts using information about the frequency and relative value of each weekday in a given month. This information is known as the "month composition". The example data set covers health insurance claims, however this technique generalizes to any other domain where monthly forecasts are required and daily data is available for calculating the month composition. Later in the post, the performance of this technique is compared to a baseline technique to show the value in a predictive forecasting scenario.

People typically seek healthcare treatment (e.g. routine check-ups, specialist treatments, etc.) during the normal business week while reserving the weekends for more interesting activities. In a given month, the total number of insurance claims may be relatively high if there are many days in the month relative to the average month. Claims could also be relatively high if there are more weekdays relative to the average month. Taking this concept a step further, some weekdays tend to produce more insurance claims than others. For example, Mondays are a relatively high volume day while Fridays are relatively low volume (though not as low as a Saturday or Sunday). All in the all, the "month composition" (i.e. the total number of days, the number of Monday, Tuesdays, etc.) is a good predictor of the expected number of insurance claims.

In a forecasting scenario, quantifying the month composition can improve the accuracy of the forecast model. To quantify month composition, first determine the total number of days and the number of occurences of Mondays, Tuesdays, Wednesdays, Thursdays, Fridays, Saturdays, and Sundays in the month. Then, quantify the "relative value" of each weekday compared to the average weekday. For example, Mondays are generally high volume so the relative value of a Monday will be higher than the average weekday. Finally, redefine the total number of days in each month by multiplying the weekday occurences by the weekday relative values and summing the results. The final summation is known as **effective days**. A recommended but optional step is to convert the effective days calculation to an index by dividing the effective day calculation by the annual average effective days. To adjust the outcome of interest (e.g. monthly claims), simply divide the outcome by the effective day index. For example, if there were 500 claims in a month and the effective day index is 1.05, the adjusted claims number would be 500/1.05 = 476.19.

By applying this calendar adjustment to the outcome of interest, the variance in the target variable caused by the month composition is removed. This adjustment combined with a population adjustment is likely to have a significantly positive effect on the final forecast model.

## R packages

When working with dates, I highly recommend  [lubridate](https://lubridate.tidyverse.org/) and of course [dplyr](https://dplyr.tidyverse.org/). For time series forecasting, I am intrigued by the [tidyverts](https://tidyverts.org/) framework. This framework offers an API for forecasting using tidy programming concepts. It has a lot of potential but is currently a bit immature. 

```{r}
library(tidyverse)
library(lubridate)
library(fable)
library(fabletools)
library(tsibble)
```

## Data

The sample data can be downloaded [here](/files/sample_claims.csv). It is a generic data set curated for the purpose of this post.

### Daily claims

The original data contains daily claim counts. In order to calculate the calendar adjustments, daily data is required.

```{r}
claims <- read_csv("sample_claims.csv")

claims 
```

### Monthly claims

Since the ultimate goal is to forecast monthly claims, the daily data is aggregated to monthly data.

```{r}
monthly_claims <- claims %>%
  group_by(Inc_Month, Inc_Year) %>%
  summarise(
    Total_Claims = sum(Claims),
    Population = mean(Population)
  ) %>%
  ungroup()

monthly_claims
```

## Quantify month composition

Quantifying month composition involves the following steps:

1. For each month in the data, count the number of occurences of each weekday

2. For each month, assign a relative weight to each weekday based on the average claims volume observed for each weekday

3. For each month, calculate effective days by multiplying the weekday occurences by the weekday relative values. Sum these products to arrive at the final effective days calculation.

### Weekday occurences per month

```{r}
weekday_counts <- claims %>%
  group_by(Inc_Month, Inc_Year, DOW_Name, DOW) %>%
  summarise(
    Total_Claims = sum(Claims),
    Occurences = n_distinct(Date)
  ) %>%
  ungroup()

weekday_counts
```

### Weekday relative values per month

Different weekdays produce different levels of claims volume. Mondays tend to produce the highest volume while Sundays tend to produce the lowest.  The boxplot below shows the distribution of claims for each weekday.

```{r}
claims %>%
  ggplot(aes(x = as.character(DOW), y = Claims)) +
  geom_boxplot() +
  labs(title = "Claims distribution by day of week",
       subtitle = "Claims volume is highest on Monday and declines over the course of a week",
       x = "Weekday") +
  theme_bw()
```

To calculate the weekday relative values, apply the following steps:

1. For each month, calculate the weekday average for each weekday by dividing the total number of claims by the number of occurences. Example: 100 claims spread across 4 Mondays would be a weekday avearge of 25.

2. Convert the weekday average to a ratio by dividing the weekday average by the average of weekday averages for the same year. 

```{r}
weekday_averages <- weekday_counts %>%
  mutate(Weekday_Average = Total_Claims/Occurences)

weekday_relatives <- weekday_averages %>%
  group_by(Inc_Month, Inc_Year) %>%
  mutate(DOW_Ratio = Weekday_Average/mean(Weekday_Average)) %>%
  ungroup() %>%
  select(Inc_Month,
         Weekday_Average,
         DOW,
         DOW_Name,
         DOW_Ratio)
```

```{r}
weekday_relatives %>%
  ggplot(aes(x = Inc_Month, y = DOW_Ratio)) +
  facet_wrap(~DOW, scales = "free_y") +
  geom_line() +
  labs(title = "Monthly weekday ratios",
       subtitle = "Notice the y-axis scales are different based on the weekday") +
  theme_bw()
```

### Calculate effective days

```{r}
effective_days <- weekday_counts %>%
  inner_join(weekday_relatives, by = c("Inc_Month", "DOW_Name")) %>%
  mutate(Effective_Days = Occurences*DOW_Ratio) %>%
  group_by(Inc_Month, Inc_Year) %>%
  summarise(
    Actual_Days = sum(Occurences),
    Effective_Days = sum(Effective_Days)
  ) %>%
  group_by(Inc_Year) %>%
  mutate(Effective_Days_Index = Effective_Days/mean(Effective_Days)) %>%
  ungroup() %>%
  select(-Inc_Year)

effective_days
```

## Calculate adjusted PMPM using month composition and population adjustments

Apply the effective days calendar adjustment *C* and a population adjustment *P* to the original monthly claims totals *Y*. The formula for doing this is $Y_{adj} = C_i\div P_i\div Y_i$, where *i* represents each month. This metric is known as adjusted PMPM (per member per month). These adjustments reduce some of the variance in the monthly insurance claims time series and make it more predictable.

```{r}
pmpm <- monthly_claims %>% 
  inner_join(effective_days, by = "Inc_Month") %>%
  mutate(Adj_PMPM = Total_Claims/Population/Effective_Days_Index) %>%
  mutate(Raw_PMPM = Total_Claims/Population) %>%
  select(Inc_Month,
         Adj_PMPM,
         Raw_PMPM,
         Total_Claims,
         Population,
         Effective_Days_Index)
```

## Compare RMSE of forecasts using adjusted and unadjusted PMPM

To test the effectiveness of a monthly claims forecast model built using adjusted PMPM, compare it to an identical model built using unadjusted PMPM. Both models are trained using a dedicated training set and evaluated using a dedicated holdout set.

### Training and holdout sets

```{r}
pmpm_ts <- pmpm %>%
  mutate(Inc_Month = yearmonth(Inc_Month)) %>%
  tsibble::as_tsibble(index = Inc_Month)

training <- pmpm_ts %>%
  filter_index(. ~ "2018-12-01")

holdout <- pmpm_ts %>%
  filter_index("2019-01-01" ~ .)
```

### Arima models

Both models are trained using an automatic ARIMA specification.

```{r}
adj_pmpm_arima <- training %>%
  model(Adj_PMPM = ARIMA(Adj_PMPM))

raw_pmpm_arima <- training %>% 
  model(Raw_PMPM = ARIMA(Raw_PMPM))
```

### Forecasts

The forecasts for both models are backtransformed to the original claims scale.

```{r}
adj_pmpm_fcast <- forecast(adj_pmpm_arima, h = 12) %>%
  mutate(Population = holdout$Population) %>%
  mutate(Effective_Days_Index = holdout$Effective_Days_Index) %>%
  mutate(Total_Claims = holdout$Total_Claims) %>%
  mutate(Forecast = .mean*Population*Effective_Days_Index) %>%
  mutate(Error = Forecast - Total_Claims)

raw_pmpm_fcast <- forecast(raw_pmpm_arima, h = 12) %>%
  mutate(Population = holdout$Population) %>%
  mutate(Effective_Days_Index = holdout$Effective_Days_Index) %>%
  mutate(Total_Claims = holdout$Total_Claims) %>%
  mutate(Forecast = .mean*Population) %>%
  mutate(Error = Forecast - Total_Claims)
```

### Errors

```{r}
tibble::tribble(
  ~Model, ~Metric, ~Value,
  "Adjusted PMPM", "RMSE", sqrt(mean(adj_pmpm_fcast$Error**2)),
  "Unadjusted PMPM", "RMSE", sqrt(mean(raw_pmpm_fcast$Error**2))
) %>%
  ggplot(aes(x = Model, y = Value)) +
  geom_col(aes(fill = Model), alpha = 0.7, width = 0.5, show.legend = F) +
  scale_fill_manual(values = c("blue", "gray50")) +
  labs(title = "Comparison of RMSE between models",
       subtitle = "Model using adjusted PMPM has lowest error on holdout set",
       y = "RMSE") +
  theme_bw()
```

## Conclusion

Month composition is quantified by determing the frequency and relative value of each weekday in a given month. When used to adjust the target variable in a forecasting scenario, it may improve accuracy as shown in this post.
